Still Missing a Cool Title

### 2006-05-31 talk/dls2006/paper.tex
Verbatimized.
Adding a bit of text, and fixing e.g. => e.g.\

Despite the constant trend in the programming world towards
portability and reusability, there are some areas in which it is still
notoriously difficult to write flexible, portable, and reasonably
efficient programs.  The implementation of virtual machines is one
such area.  Building implementations of general programming languages,
in particular highly dynamic ones, using a classic direct coding
approach, is typically a long-winded effort and produces a result that
is quite [XXX quite could be removed here?] tailored to a specific
platform and where architectural decisions (e.g. about GC) are spread
across the code in a pervasive and invasive way.

@@

\item We perform type inference on the control flow graphs.  At this
      stage, types inferred are part of the type system which is the
      very definition of the RPython sub-language: they are roughly a
      subset of Python's built-in types, with some more precision to
      describe e.g. the items stored in container types.
      Occasionally, a single input function can produce several
      specialized versions, i.e. several similar but differently typed
      graphs.  This type inference process is described in more
      details in section \ref{typeinference}.
\end{enumerate}
 %     
At the end of the front-end analysis, the input RPython program is
represented as a forest of flow graphs with typed variables.  Following
this analysis are a number of transformation steps.  Each transformation
step modifies the graphs in-place, by altering their structure and/or
the operations they contain.  Each step inputs graphs typed in one type
system and leaves them typed in a possibly different type system, as we
will describe in the sequel.  Finally, a back-end turns the resulting
graphs into code suitable for the target environment, e.g. C source code
ready to be compiled.

@@

The other, and better, alternative is an exact GC, coupled with a
transformation, the \textit{GC transformer}.  It inputs C-level-typed graphs
and replaces all \texttt{malloc} operations with calls to a garbage
collector's innards.  It can inspect all the graphs to discover the
\texttt{struct} types in use by the program, and assign a unique type id to
each of them.  These type ids are collected in internal tables that
describe the layout of the structures, e.g. their sizes and the location
of the pointer fields.

We have implemented other transformations as well, e.g. performing
various optimizations, or turning the whole code into a
continuation-passing style (CPS) [XXX I'm not sure our transformation
can be classified as classical CPS, although there are known similar techniques but the terminology is quite confused] that allows us to use coroutines
without giving up the ability to generate fully ANSI C code.  (This will
be the subject of another paper.)  [XXX mention exception transformer too]

@@

Note that such system code cannot typically be expressed as normal
RPython functions, because it corresponds to primitive operations at
that level.  As an aside, let us remark that the number of primitive
operations at RPython level is, comparatively speaking, quite large:
all list and dictionary operations, instance and class attribute
accesses, many string processing methods, a good subset of all Python
built-in functions...  Compared to other approaches [e.g. Squeak], we
do not try to minimize the number of primitives -- at least not at the
source level.  It is fine to have many primitives at any high enough
level, because they can all be implemented at the next lower level in
a way that makes sense to that level.  The key reason why this is not
burdensome is that the lower level implementations are also written in
Python -- with the only difference that they use (and have to be
typeable in) the lower-level type system\footnote{This is not strictly
true: the type systems are even allowed to co-exist in the same
function.  The operations involving higher-level type systems are
turned into lower-level operations by the previous transformations in
the chain, which leave the already-low-level operations untouched.}.

@@

The RPython level is a subset of Python, so the types mostly follow
Python types, and the instances of these types are instances in the
normal Python sense; e.g. whereas Python has only got a single type
\texttt{list}, RPython has a parametric type \texttt{list(T)} for every RPython
type \texttt{T}, but instances of \texttt{list(T)} are just those Python lists
whose items are all instances of \texttt{T}.

@@

Naturally, simply propagating types forward requires the use of a fixed
point algorithm in the presence of loops in the flow graphs or in the
inter-procedural call graph.  Indeed, we flow types forward from the
beginning of the entry point function into each basic block, operation
after operation, and follow all calls recursively.  During this process,
each variable along the way gets a type.  In various cases, e.g. when we
close a loop, the previously assigned types can be found to be too
restrictive.  In this case, we generalise them to allow for a larger set
of possible run-time values, and schedule the block where they appear
for reflowing.  The more general type can generalise the types of the
results of the variables in the block, which in turn can generalise the
types that flow into the following blocks, and so on.  This process
continues until a fixed point is reached.

@@

It is outside the scope of the present paper to describe the type
inference engine and the rules more formally.  The difficult points
involve mutable containers -- e.g. initially empty list that are filled
somewhere else -- and the discovery of instance attributes -- in Python,
classes do not declare upfront a fixed set of attributes for their
instances, let alone their types.  Both these examples require
sophisticated reflowing techniques that invalidate existing types in
already-annotated basic blocks, to account for the influence of more
general types coming indirectly from a possibly distant part of the
program.  The reader is referred to [D] for more information.

@@

The worst case behaviors that can appear in the model described above
involve the lattice of Pbcs, involving variables that could contain
e.g.  one function object among many.  An example of such behavior is
code manipulating a table of function objects: when an item is read
out of the table, its type is a large Pbc set: $Pbc(\{f1, f2, f3,
\ldots\})$.  But in this example, the whole set is available at once,
and not built incrementally by successive discoveries.  This seems to
be often the case in practice: it is not very common for programs to
manipulate objects that belong to a large but finite family -- and when
they do, the whole family tends to be available early on, requiring
few reflowing.

@@

Of course, this would be pointless if the annotation did not give
precise enough information for our needs.  We must describe a detail of
the abstract interpretation engine that is critical for precision: the
propagation of conditional types.  Consider the following source code
fragment::

    if isinstance(x, MyClass):
        f(x)
    else:
        g(x)

>>>>

Despite the constant trend in the programming world towards
portability and reusability, there are some areas in which it is still
notoriously difficult to write flexible, portable, and reasonably
efficient programs.  The implementation of virtual machines is one
such area.  Building implementations of general programming languages,
in particular highly dynamic ones, using a classic direct coding
approach, is typically a long-winded effort and produces a result that
is tailored to a specific
platform and where architectural decisions (e.g.\ about GC) are spread
across the code in a pervasive and invasive way.

@@

\item We perform type inference on the control flow graphs.  At this
      stage, types inferred are part of the type system which is the
      very definition of the RPython sub-language: they are roughly a
      subset of Python's built-in types, with some more precision to
      describe e.g.\ the items stored in container types.
      Occasionally, a single input function can produce several
      specialized versions, i.e. several similar but differently typed
      graphs.  This type inference process is described in more
      details in section \ref{typeinference}.
\end{enumerate}
 %     
At the end of the front-end analysis, the input RPython program is
represented as a forest of flow graphs with typed variables.  Following
this analysis are a number of transformation steps.  Each transformation
step modifies the graphs in-place, by altering their structure and/or
the operations they contain.  Each step inputs graphs typed in one type
system and leaves them typed in a possibly different type system, as we
will describe in the sequel.  Finally, a back-end turns the resulting
graphs into code suitable for the target environment, e.g.\ C source code
ready to be compiled.

@@

The other, and better, alternative is an exact GC, coupled with a
transformation, the \textit{GC transformer}.  It inputs C-level-typed graphs
and replaces all \texttt{malloc} operations with calls to a garbage
collector's innards.  It can inspect all the graphs to discover the
\texttt{struct} types in use by the program, and assign a unique type id to
each of them.  These type ids are collected in internal tables that
describe the layout of the structures, e.g.\ their sizes and the location
of the pointer fields.

We have implemented other transformations as well, e.g.\ performing
various optimizations, or turning the whole code into a style that
allows us to use coroutines (still in ANSI C: it is a variant of
continuation-passing that will be the subject of another paper.)
Another example is the exception transformer, which transforms graphs
that still contain implicit exception handling into a form suitable for
C (currently based on a global flag to signal the presence of an
exception, set and checked around selected function calls).

@@

Note that such system code cannot typically be expressed as normal
RPython functions, because it corresponds to primitive operations at
that level.  As an aside, let us remark that the number of primitive
operations at RPython level is, comparatively speaking, quite large:
all list and dictionary operations, instance and class attribute
accesses, many string processing methods, a good subset of all Python
built-in functions...  Compared to other approaches [e.g.\ Squeak], we
do not try to minimize the number of primitives -- at least not at the
source level.  It is fine to have many primitives at any high enough
level, because they can all be implemented at the next lower level in
a way that makes sense to that level.  The key reason why this is not
burdensome is that the lower level implementations are also written in
Python -- with the only difference that they use (and have to be
typeable in) the lower-level type system\footnote{This is not strictly
true: the type systems are even allowed to co-exist in the same
function.  The operations involving higher-level type systems are
turned into lower-level operations by the previous transformations in
the chain, which leave the already-low-level operations untouched.}.

@@

The RPython level is a subset of Python, so the types mostly follow
Python types, and the instances of these types are instances in the
normal Python sense; e.g.\ whereas Python has only got a single type
\texttt{list}, RPython has a parametric type \texttt{list(T)} for every RPython
type \texttt{T}, but instances of \texttt{list(T)} are just those Python lists
whose items are all instances of \texttt{T}.

@@

Naturally, simply propagating types forward requires the use of a fixed
point algorithm in the presence of loops in the flow graphs or in the
inter-procedural call graph.  Indeed, we flow types forward from the
beginning of the entry point function into each basic block, operation
after operation, and follow all calls recursively.  During this process,
each variable along the way gets a type.  In various cases, e.g.\ when we
close a loop, the previously assigned types can be found to be too
restrictive.  In this case, we generalise them to allow for a larger set
of possible run-time values, and schedule the block where they appear
for reflowing.  The more general type can generalise the types of the
results of the variables in the block, which in turn can generalise the
types that flow into the following blocks, and so on.  This process
continues until a fixed point is reached.

@@

It is outside the scope of the present paper to describe the type
inference engine and the rules more formally.  The difficult points
involve mutable containers -- e.g.\ initially empty list that are filled
somewhere else -- and the discovery of instance attributes -- in Python,
classes do not declare upfront a fixed set of attributes for their
instances, let alone their types.  Both these examples require
sophisticated reflowing techniques that invalidate existing types in
already-annotated basic blocks, to account for the influence of more
general types coming indirectly from a possibly distant part of the
program.  The reader is referred to [D] for more information.

@@

The worst case behaviors that can appear in the model described above
involve the lattice of Pbcs, involving variables that could contain
e.g.\ one function object among many.  An example of such behavior is
code manipulating a table of function objects: when an item is read
out of the table, its type is a large Pbc set: $Pbc(\{f1, f2, f3,
\ldots\})$.  But in this example, the whole set is available at once,
and not built incrementally by successive discoveries.  This seems to
be often the case in practice: it is not very common for programs to
manipulate objects that belong to a large but finite family -- and when
they do, the whole family tends to be available early on, requiring
few reflowing.

@@

Of course, this would be pointless if the annotation did not give
precise enough information for our needs.  We must describe a detail of
the abstract interpretation engine that is critical for precision: the
propagation of conditional types.  Consider the following source code
fragment:

\begin{verbatim}
    if isinstance(x, MyClass):
        f(x)
    else:
        g(x)
\end{verbatim}
